---
title: "Computer Practical 1"
subtitle: "Statistics 2"
author: "Dom Hutchinson"
output:
  pdf_document: 
     fig_width: 6
     fig_height: 4
  html_notebook: default
  html_document: default
  word_document: default
---

# 2 Binomial Maximum Likelihood Estimators

Let $Y\sim\text{Binomial}(n,p)$.
The maximum likelihood estimate for $p$ is $\hat{p}(Y)=\frac{Y}{n}$. $\hat{p}$ is an unbiased estimator.

## Question 1
Let $X_1,\dots,X_n\overset{\text{iid}}{\sim}\text{Bernoulli}(p)$. Then
$$
\begin{array}{rcl}
\text{Var}(\hat{p})&=&\text{Var}\left(\frac{1}{n}\sum\limits_{i=1}^nX_i\right)\\
&=&\dfrac{1}{n^2}\text{Var}\left(\sum\limits_{i=1}^nX_i\right)\\
&=&\dfrac{1}{n^2}.n\text{Var}\left(X_1\right)\\
&=&\dfrac{1}{n}p(1-p)
\end{array}
$$

## Question 2

```{r}
n<-13; p<-.31; sample_size<-25; trials<-1000
var_hat=(1/n)*p*(1-p)
var_hat

phat<-function(Y) {
  Y/n
}

x_values<-rbinom(n=sample_size*trials,size=n,prob=p)
x_samples<-matrix(x_values,nrow=sample_size)
p_hat.samples<-apply(x_samples,1,phat)
p_hat.sigma2=apply(p_hat.samples,1,var)
hist(p_hat.sigma2,breaks=20)
abline(v=var_hat,col="red")
```

# 3 Clinic Data

```{r}
year.data<-read.csv("AssessedProblemSheet1S2_files/year_data.csv")
knitr::kable(year.data)
```

Let $Y_i\sim\text{Binomial}(n_i,p_i)$ model the number of deaths in clinic $i$ where $n_i$ is the total number of births in clinic $i$ & $p_i$ is the mortality rate for clinic $i$. Assume $Y_1$ & $Y_2$ are independent.

```{r}
n1 <- sum(year.data[year.data$clinic==1,]$births) # number of births in clinic 1
y1 <- sum(year.data[year.data$clinic==1,]$deaths) # number of deaths in clinic 1
cat("Number of births in clinic 1:",prettyNum(n1,big.mark=","),"\nNumber of deaths in clinic 1:",prettyNum(y1,big.mark=","),"\n\n")

n2 <- sum(year.data[year.data$clinic==2,]$births) # number of births in clinic 2
y2 <- sum(year.data[year.data$clinic==2,]$deaths) # number of deaths in clinic 2
cat("Number of births in clinic 2:",prettyNum(n2,big.mark=","),"\nNumber of deaths in clinic 2:",prettyNum(y2,big.mark=","),"\n")
```

## Question 3

```{r}
p1_hat=y1/n1
p2_hat=y2/n2

cat("p1_hat:",p1_hat,"\np2_hat:",p2_hat)
```

## Quetion 4
Assume that $p=p_1=p_2$ and define $W:=\hat{p}_1(Y_1)-\hat{p}_2(Y_2)$. Then
$$
\begin{array}{rcl}
\mathbb{E}(W)&=&\mathbb{E}(\hat{p}_1(Y_1)-\hat{p}_2(Y_2))\\
&=&\mathbb{E}(\hat{p}_1(Y_1))-\mathbb{E}(\hat{p}_2(Y_2))\\
&=&p_1-p_2\\
&=&p-p\\
&=&0\\
\text{Var}(W)&=&\text{Var}(\hat{p}_1(Y_1)-\hat{p}_2(Y_2))\\
&=&\text{Var}(\hat{p}_1(Y_1))+\text{Var}(\hat{p}_2(Y_2))\\
&=&\frac{1}{n_1}p_1(1-p_1)+\frac{1}{n_2}p_2(1-p_2)\\
&=&\frac{1}{n_1}p(1-p)+\frac{1}{n_2}p(1-p)\\
&=&\frac{n_1+n_2}{n_1n_2}p(1-p)
\end{array}
$$

## Question 5
Suppose $p=p_1=p_2$. We have $\hat{p}=\frac{`r y1`+`r y2`}{`r n1`+`r n2`}=\frac{`r y1+y2`}{`r n1+n2`}=`r p_hat<-(y1+y2)/(n1+n2); p_hat`$.
$$
\begin{array}{rcl}
\mathbb{P}(|W-\mu_W|\geq\hat{p}_1(y_1)-\hat{p}_1(y_2))&=&\mathbb{P}(|W-0|\geq`r p1_hat`-`r p2_hat`)\\
&=&\mathbb{P}(|W|\geq`r p1_hat-p2_hat`)\\
&\leq&\dfrac{\sigma^2_W}{`r p1_hat-p2_hat`^2}\text{ by Chebyshev's Inequality}\\
&=&\dfrac{1}{`r p1_hat-p2_hat`^2}\times\dfrac{n_1+n_2}{n_1n_2}\hat{p}(1-\hat{p})\\
&=&\dfrac{1}{`r p1_hat-p2_hat`^2}\times\dfrac{`r n1`+`r n2`}{`r n1`\times`r n2`}\times`r p_hat`\times`r 1-p_hat`\\
&=&`r (1/(p1_hat-p2_hat)^2)*((n1+n2)/(n1*n2))*p_hat*(1-p_hat)`
\end{array}
$$
Thus it is very unlikely to observe these two mortality rates, assuming the underlying rate is the same.

# 4 Intervention: Chlorine Hand Washing

```{r}
month.data<-read.csv("AssessedProblemSheet1S2_files/month_data.csv")
month.data<-month.data[!is.na(month.data$births),]

month.data$rate<-month.data$deaths/month.data$births
month.data$date<-as.Date(month.data$date)

intervention.date<-as.Date("1847-05-15")

plot(month.data$date, month.data$rate, pch=20, main="Mortality Rate by Month", xlab="Date (Red line indicates start of intervention preiod)", ylab="Rate")
abline(v=intervention.date, col="red")
```

```{r}
before.intervention<-month.data[month.data$date< intervention.date,]
n1<-sum(before.intervention$births)
y1<-sum(before.intervention$deaths)
cat("Number of births before intervention:",prettyNum(n1,big.mark=","),"\nNumber of deaths before intervention:",prettyNum(y1,big.mark=","),"\n\n")

after.intervention <-month.data[month.data$date>=intervention.date,]
n2<-sum(after.intervention$births)
y2<-sum(after.intervention$deaths)
cat("Number of births after intervention:",prettyNum(n2,big.mark=","),"\nNumber of deaths after intervention:",prettyNum(y2,big.mark=","))
```

## Question 6
```{r}
p1_hat<-y1/n1
p2_hat<-y2/n2

cat("p1_hat:",p1_hat,"\np2_hat:",p2_hat)
```
Define random variable $W=\hat{p}_1(Y_1)-\hat{p}_2(Y_2)$.

Suppose $p=p_1=p_2$. We have $\hat{p}=\frac{`r y1`+`r y2`}{`r n1`+`r n2`}=\frac{`r y1+y2`}{`r n1+n2`}=`r p_hat<-(y1+y2)/(n1+n2); p_hat`$.
$$
\begin{array}{rcl}
\mathbb{P}(|W-\mu_W|\geq\hat{p}_1(y_1)-\hat{p}_1(y_2))&=&\mathbb{P}(|W-0|\geq`r p1_hat`-`r p2_hat`)\\
&=&\mathbb{P}(|W|\geq`r p1_hat-p2_hat`)\\
&\leq&\dfrac{\sigma^2_W}{`r p1_hat-p2_hat`^2}\text{ by Chebyshev's Inequality}\\
&=&\dfrac{1}{`r p1_hat-p2_hat`^2}\times\dfrac{n_1+n_2}{n_1n_2}\hat{p}(1-\hat{p})\\
&=&\dfrac{1}{`r p1_hat-p2_hat`^2}\times\dfrac{`r n1`+`r n2`}{`r n1`\times`r n2`}\times`r p_hat`\times`r 1-p_hat`\\
&=&`r (1/(p1_hat-p2_hat)^2)*((n1+n2)/(n1*n2))*p_hat*(1-p_hat)`
\end{array}
$$
Thus it is very unlikely to observe these two mortality rates, assuming the underlying rate is the same.

# 5 A First Logistic Regression

```{r}
x1<-c(1,0)
x2<-c(1,1)

sigma<-function(z) {
  1/(1+exp(-z))
}
```

## Question 7
