---
title: "Computer Practical 1"
subtitle: "Statistics 2"
author: "Dom Hutchinson"
output:
  pdf_document: 
     fig_width: 6
     fig_height: 4
  html_notebook: default
  html_document: default
  word_document: default
---

# 2 Binomial Maximum Likelihood Estimators

Let $Y\sim\text{Binomial}(n,p)$.
The maximum likelihood estimate for $p$ is $\hat{p}(Y)=\frac{Y}{n}$. $\hat{p}$ is an unbiased estimator.

## Question 1
Let $X_1,\dots,X_n\overset{\text{iid}}{\sim}\text{Bernoulli}(p)$. Then
$$
\begin{array}{rcl}
\text{Var}(\hat{p})&=&\text{Var}\left(\dfrac{Y}{n}\right)\\
&=&\text{Var}\left(\dfrac{1}{n}\sum\limits_{i=1}^nX_i\right)\\
&=&\dfrac{1}{n^2}\text{Var}\left(\sum\limits_{i=1}^nX_i\right)\\
&=&\dfrac{1}{n^2}.n\text{Var}\left(X_1\right)\\
&=&\dfrac{1}{n}p(1-p)
\end{array}
$$

## Question 2

```{r}
n<-13; p<-.31
sample_size<-100; trials=1000
var_hat=(1/n)*p*(1-p)
cat("var_hat:",var_hat,"\n")

phat<-function(Y) {
  Y/n
}

x_values<-rbinom(n=sample_size*trials,size=n,prob=p)
x_samples<-matrix(x_values,nrow=sample_size)
p_hat.samples<-apply(x_samples,1,phat)
p_hat.sigma2=apply(p_hat.samples,1,var)
cat("observered_var:",mean(p_hat.sigma2))
hist(p_hat.sigma2,breaks=10,xlab="Variance (Red line=var_hat)")
abline(v=var_hat,col="red")

```

# 3 Clinic Data

```{r}
year.data<-read.csv("data/year_data.csv")
knitr::kable(year.data)
```

Let $Y_i\sim\text{Binomial}(n_i,p_i)$ model the number of deaths in clinic $i$ where $n_i$ is the total number of births in clinic $i$ & $p_i$ is the mortality rate for clinic $i$. Assume $Y_1$ & $Y_2$ are independent.

```{r}
n1 <- sum(year.data[year.data$clinic==1,]$births) # number of births in clinic 1
y1 <- sum(year.data[year.data$clinic==1,]$deaths) # number of deaths in clinic 1
cat("Number of births in clinic 1:",prettyNum(n1,big.mark=","),"\nNumber of deaths in clinic 1:",prettyNum(y1,big.mark=","),"\n\n")

n2 <- sum(year.data[year.data$clinic==2,]$births) # number of births in clinic 2
y2 <- sum(year.data[year.data$clinic==2,]$deaths) # number of deaths in clinic 2
cat("Number of births in clinic 2:",prettyNum(n2,big.mark=","),"\nNumber of deaths in clinic 2:",prettyNum(y2,big.mark=","),"\n")
```

## Question 3

```{r}
p1_hat=y1/n1
p2_hat=y2/n2

cat("p1_hat:",p1_hat,"\np2_hat:",p2_hat)
```

## Quetion 4
Assume that $p=p_1=p_2$ and define $W:=\hat{p}_1(Y_1)-\hat{p}_2(Y_2)$. Then
$$
\begin{array}{rcl}
\mathbb{E}(W)&=&\mathbb{E}(\hat{p}_1(Y_1)-\hat{p}_2(Y_2))\\
&=&\mathbb{E}(\hat{p}_1(Y_1))-\mathbb{E}(\hat{p}_2(Y_2))\\
&=&p_1-p_2\\
&=&p-p\\
&=& 0\\
\text{Var}(W)&=&\text{Var}(\hat{p}_1(Y_1)-\hat{p}_2(Y_2))\\
&=&\text{Var}(\hat{p}_1(Y_1))+\text{Var}(\hat{p}_2(Y_2))\\
&=&\frac{1}{n_1}p_1(1-p_1)+\frac{1}{n_2}p_2(1-p_2)\\
&=&\frac{1}{n_1}p(1-p)+\frac{1}{n_2}p(1-p)\\
&=&\frac{n_1+n_2}{n_1n_2}p(1-p)
\end{array}
$$

## Question 5
Suppose $p=p_1=p_2$. We have $\hat{p}=\frac{`r y1`+`r y2`}{`r n1`+`r n2`}=\frac{`r y1+y2`}{`r n1+n2`}=`r p_hat<-(y1+y2)/(n1+n2); p_hat`$.
$$
\begin{array}{rcl}
\mathbb{P}(|W-\mu_W|\geq\hat{p}_1(y_1)-\hat{p}_1(y_2))&=&\mathbb{P}(|W- |\geq`r p1_hat`-`r p2_hat`)\\
&=&\mathbb{P}(|W|\geq`r p1_hat-p2_hat`)\\
&\leq&\dfrac{\sigma^2_W}{`r p1_hat-p2_hat`^2}\text{ by Chebyshev's Inequality}\\
&=&\dfrac{1}{`r p1_hat-p2_hat`^2}\times\dfrac{n_1+n_2}{n_1n_2}\hat{p}(1-\hat{p})\\
&=&\dfrac{1}{`r p1_hat-p2_hat`^2}\times\dfrac{`r n1`+`r n2`}{`r n1`\times`r n2`}\times`r p_hat`\times`r 1-p_hat`\\
&=&`r (1/(p1_hat-p2_hat)^2)*((n1+n2)/(n1*n2))*p_hat*(1-p_hat)`
\end{array}
$$
Thus it is very unlikely to observe these two mortality rates, assuming the underlying rate is the same.

# 4 Intervention: Chlorine Hand Washing

```{r}
month.data<-read.csv("data/month_data.csv")
month.data<-month.data[!is.na(month.data$births),]

month.data$rate<-month.data$deaths/month.data$births
month.data$date<-as.Date(month.data$date)

intervention.date<-as.Date("1847- 5-15")

plot(month.data$date, month.data$rate, pch=20, main="Mortality Rate by Month",
     xlab="Date (Red line indicates start of intervention preiod)", ylab="Rate")
abline(v=intervention.date, col="red")
```

```{r}
before.intervention<-month.data[month.data$date< intervention.date,]
n1<-sum(before.intervention$births)
y1<-sum(before.intervention$deaths)
cat("Number of births before intervention:",prettyNum(n1,big.mark=","),
    "\nNumber of deaths before intervention:",prettyNum(y1,big.mark=","),"\n\n")

after.intervention <-month.data[month.data$date>=intervention.date,]
n2<-sum(after.intervention$births)
y2<-sum(after.intervention$deaths)
cat("Number of births after intervention: ",prettyNum(n2,big.mark=","),
    "\nNumber of deaths after intervention: ",prettyNum(y2,big.mark=","))
```

## Question 6
```{r}
p1_hat<-y1/n1
p2_hat<-y2/n2

cat("p1_hat:",p1_hat,"\np2_hat:",p2_hat)
```
Define random variable $W=\hat{p}_1(Y_1)-\hat{p}_2(Y_2)$.

Suppose $p=p_1=p_2$. We have $\hat{p}=\frac{y_1+y_2}{n_1+n_2}-\frac{`r y1`+`r y2`}{`r n1`+`r n2`}=\frac{`r y1+y2`}{`r n1+n2`}=`r p_hat<-(y1+y2)/(n1+n2); p_hat`$.
$$
\begin{array}{rcl}
\mathbb{P}(|W-\mu_W|\geq\hat{p}_1(y_1)-\hat{p}_1(y_2))&=&\mathbb{P}(|W-0|\geq`r p1_hat`-`r p2_hat`)\\
&=&\mathbb{P}(|W|\geq`r p1_hat-p2_hat`)\\
&\leq&\dfrac{\sigma^2_W}{`r p1_hat-p2_hat`^2}\text{ by Chebyshev's Inequality}\\
&=&\dfrac{1}{`r p1_hat-p2_hat`^2}\times\dfrac{n_1+n_2}{n_1n_2}\hat{p}(1-\hat{p})\\
&=&\dfrac{1}{`r p1_hat-p2_hat`^2}\times\dfrac{`r n1`+`r n2`}{`r n1`\times`r n2`}\times`r p_hat`\times`r 1-p_hat`\\
&=&`r (1/(p1_hat-p2_hat)^2)*((n1+n2)/(n1*n2))*p_hat*(1-p_hat)`
\end{array}
$$
Thus. it is very unlikely to observe these two mortality rates, assuming the underlying rate is the same.

# 5 A First Logistic Regression

```{r}
x1<-c(1,0)
x2<-c(1,1)

sigma<-function(z) {
  1/(1+exp(-z))
}
```

## Question 7
$$
\begin{array}{rrcl}
&L(\theta)&\propto&\prod_{i=1}^2f_{Y_i}(y_i;n_i,x_i,\theta)\\
&&=&f(y_1;n_1,x_1,\theta)f(y_2;n_2,x_2,\theta)\\
&&=&{n_1\choose y_1}g(x_1,\theta)^{y_1}(1-g(x_1,\theta))^{n_1-y_1}{n_2\choose y_2}g(x_2,\theta)^{y_2}(1-g(x_2,\theta))^{n_2-y_2}\\
&&=&{n_1\choose y_1}\sigma(\theta_1)^{y_1}(1-\sigma(\theta_1))^{n_1-y_1}{n_2\choose y_2}\sigma(\theta_1+\theta_2)^{y_2}(1-\sigma(\theta_1+\theta_2))^{n_2-y_2}\\
&&\propto&\sigma(\theta_1)^{y_1}(1-\sigma(\theta_1))^{n_1-y_1}\sigma(\theta_1+\theta_2)^{y_2}(1-\sigma(\theta_1+\theta_2))^{n_2-y_2}\\
\implies&\ell(\theta)&=&c+y_1\ln(\sigma(\theta_1))+(n_1-y_1)\ln(1-\sigma(\theta_1))\\
&&+&y_2\ln(\sigma(\theta_1+\theta_2))+(n_2-y_2)\ln(1-\sigma(\theta_1+\theta_2))\\\\
&\sigma(\theta_1)&=&\dfrac{1}{1+e^{-\theta_1}}\\
\implies&\ln(\sigma(\theta_1))&=&-\ln(1+e^{-\theta_1})\\
\&&\ln(1-\sigma(\theta_1))&=&\ln\left(\frac{e^{-\theta_1}}{1+e^{-\theta_1}}\right)\\
&&=&-\theta_1-\ln(1+e^{-\theta_1})\\\\
&\sigma(\theta_1+\theta_2)&=&\dfrac{1}{1+e^{-(\theta_1+\theta_2)}}\\
\implies&\ln(\sigma(\theta_1+\theta_2))&=&-\ln(1+e^{-(\theta_1+\theta_2)})\\
\&&\ln(1-\sigma(\theta_1+\theta_2))&=&\ln\left(\frac{e^{-(\theta_1+\theta_2)}}{1+e^{-(\theta_1+\theta_2)}}\right)\\
&&=&-(\theta_1+\theta_2)-\ln(1+e^{-(\theta_1+\theta_2)})\\\\
\implies&\ell(\theta)&=&c-y_1\ln(1+e^{\theta_1})-(n_1-y_1)(\theta_1+\ln(1+e^{-\theta_1}))\\
&&-&y_2\ln(1+e^{-(\theta_1+\theta_2)})-(n_2-y_2)(\theta_1+\theta_2+\ln(1+e^{-(\theta_1+\theta_2)}))\\
&&=&c-n_1\ln(1+e^{-\theta_1})-n_2\ln(1+e^{-(\theta_1+\theta_2)})-\theta_1(n_1-y_1)-(\theta_1+\theta_2)(n_2-y_2)\\
&&=&c-`r n1`\ln(1+e^{-\theta_1})-`r n2`\ln(1+e^{-(\theta_1+\theta_2)})-\theta_1(`r n1`-`r y1`)-(\theta_1+\theta_2)(`r n2`-`r y1`)\\
&&=&c-`r n1`\ln(1+e^{-\theta_1})-`r n2`\ln(1+e^{-(\theta_1+\theta_2)})-`r n1-y1`\theta_1-`r n2-y1`(\theta_1+\theta_2)\\
&&=&c-`r n1`\ln(1+e^{-\theta_1})-`r n2`\ln(1+e^{-(\theta_1+\theta_2)})-`r n1-y1+(n2-y1)`\theta_1-`r n2-y1`\theta_2
\end{array}\\
$$
```{r}
ell<-function(theta) {
  l<--19571*log(1+exp(-theta[1]))-6595*log(1+exp(-theta[2]))-22046*theta[1]-4535*theta[2]
  -l # In order to find maximum
}

result<-optim(c(.5,.5), ell)
cat("theta_hat:(",result$par[1],",",result$par[2],")")
```
This value of $\hat{\theta}_1$ shows that the probability of mortality before intervention was very low, which is corroborated by $\hat{p}_1$ being low. This value of $\hat{\theta}_2$ shows that the probability of mortality decreased after intervention, this is corroborated by $\hat{p}_2<\hat{p}_1$.